Estimation of Spatiotemporal Integro-Difference Equations for Dynamical Neural Field Models

### 2009-11-16 ltx/BrainIDE.tex
Fixed notational things in model derivation and fixed some typos

\section{Neural Field Model}
In this section we describe a variant of the  Wilson and Cowan neural field model~\cite{Wilson1973} that will be used to create our connectivity estimator. This model has shown to be descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations and epileptic behavior. It is also capable of generating complex patterns of activity such as Turing patterns, spirals and traveling oscillations. The general form of the model is described here. The post-synaptic potentials generated at a neuronal population $s$ by action potentials arriving from all other connected populations at position $r$ can be described by
\begin{equation}\label{Eq1}
v\left( {s,t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)J\left( {s,t'} \right)dt'}.
\end{equation}
The post-synaptic response kernel $h(t)$ is described by
\begin{equation}\label{Eq2}
h(t) = \left\{ {\begin{array}{*{20}{c}}
   {\exp ( - {t \mathord{\left/
 {\vphantom {t \tau }} \right.
 \kern-\nulldelimiterspace} \tau })} & {t \ge 0}  \\
   0 & {t < 0}  \\
\end{array}} \right.,
\end{equation}
where $\tau$ is the synaptic time constant and typically has values from 10-20~ms (\textcolor[rgb]{1.00,0.00,0.00}{refs}). The synaptic response kernel can be more elaborate, using a rise and fall time constant, however this simple form serves our purposes. Equation~\ref{Eq1} relates average firing rate of a presynaptic neural population to the mean membrane potential of the post-synaptic neurons. The mean firing rate, $J$, is convolved with the response kernel to give membrane potential since the firing rate is assumed to follow a (homogeneous?) Poisson process. Nonlocal interactions between cortical populations are described by	
\begin{equation}\label{Eq3}
J\left( {s,t} \right) = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right)
\end{equation}
where $\bar t = t - \frac{{\left| {s - r} \right|}}{d}$, $P$ is a mean pulse rate of population at $y$, $W$ is the connectivity kernel, $E$ is an external input, $\Omega$ is the spatial domain and $d$ is the propagation velocity from $r$ to $s$.

The average firing rate of the presynaptic neurons rate to the pre-synaptic membrane potential by the Heaviside step function
\begin{equation}\label{Eq4}
P\left( {v\left( {r,\bar t} \right)} \right) = {r_0}S\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right)
\end{equation}
\begin{equation}\label{Eq4.1}
S\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right) = \left\{ {\begin{array}{*{20}{c}}
   1 & {v\left( {r,\bar t} \right) \ge {v_0}\left( r \right)}  \\
   0 & {otherwise}  \\
\end{array}} \right.
\end{equation}
Even though $v_0$ is a constant in this research we show express it as a field allowing for inhomogeneities and for clarity and generality in constructing the state-space representation in the next section. The parameters in equation~\ref{Eq4} are... $v_0$ is typically 6~mV, $r_0$ is 5~$s^{-1}$ (from Wendling and David papers). By combining equations~\ref{Eq1} and \ref{Eq3} we get the spatiotemporal model
\begin{equation}\label{Eq5}
v\left(s,t\right) = \int\limits_{-\infty}^t\int\limits_\Omega  h\left(t - t'\right)W\left(s,r\right)P\left( v\left( r,\bar t \right)\right)dr dt'+E\left(s,t'\right)
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{Eq6}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{Eq7}
L\left( {\frac{\partial }{{dt}}} \right)v\left( {s,t} \right) = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right),
\end{equation}
where $L$ is a polynomial of order $n$ with constant coefficients. This provides an $n^{th}$ order model. Most studies considering this model use first or second order time derivatives. We shall use the first order model giving
\begin{equation}\label{Eq8}
\frac{{dv\left( {s,t} \right)}}{{dt}} + \frac{{v\left( {s,t} \right)}}{\tau } = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right).
\end{equation}
\subsection{Alternate Model Description I}
An alternate and equivalent way to remove the temporal integration is through the Laplace transform. We start by considering a discrete neural mass $i$ with inputs with zero delay from another discrete mass $j$ and
\begin{equation}\label{Eq9.1}
{v_i}\left( t \right) = \int_{ - \infty }^t {{h_j}(t - t'){w_{ij}}P\left( {{v_j}\left( t \right)} \right)dt'},
\end{equation}
where $w_{ij}$ can be considered the synaptic efficiency from $j$ to $i$. Now we take the Laplace transform yielding
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  {V_i}\left( s \right) &=& {H_j}\left( s \right){w_{ij}}P\left( {{V_j}\left( s \right)} \right) \nonumber \\
   &=& {\left( {s + \frac{1}{\tau }} \right)^{ - 1}}{w_{ij}}P\left( {{V_j}\left( s \right)} \right)  \\
  \Rightarrow s{V_i}\left( s \right) + \frac{1}{\tau }{V_i}\left( s \right) &=& {w_{ij}}P\left( {{V_j}\left( s \right)} \right).
\end{eqnarray}
Now taking the inverse Laplace transform we get
\begin{equation}\label{9.3}
\frac{{d{v_i}\left( t \right)}}{{dt}} + \frac{1}{\tau }{v_i}\left( t \right) = {w_{ij}}P\left( {{v_j}\left( t \right)} \right).
\end{equation}
Next we convert the model from discrete neural masses to a continuum by
\begin{equation}\label{Eq9.4}
\frac{{dv\left( {s,t} \right)}}{{dt}} + \frac{1}{\tau }{v\left( {s,t} \right)} = \int_\Omega  {W\left( {s,r} \right)P\left( {v\left( {r,\bar t} \right)} \right)dr}  + E\left( {s,t} \right).
\end{equation}
The neural field model is popular due to being parsimonious yet having a strong link with the underlying physiology.

@@

An alternative description of the model is as follows. The model can be thought of as a continuum of neural masses. To illustrate this idea we shall first describe the interaction of discrete neural masses as seen in Figure~\ref{NeuralMassModel}. Assuming post-synaptic potentials sum linearly, the membrane potential at the soma of neuron $i$ is described as
\begin{equation}\label{Eq1.1}
{v_i}\left( t \right) = \sum\nolimits_{j,k} {{w_{ij}}{h_{j}}\left( t - {t_k}\right)} H\left( {t - {t_k}} \right),
\end{equation}
where $w_{ij}$ is the synaptic weight from neuron $j$ to $i$, $h(t)$ is the synaptic response kernel, $H$ is the Heaviside step function and action potential arrive at neuron $i$ at times $t_k$. Alternatively the membrane potential can described by the convolution of synaptic response kernel with the pre-synaptic firing rate $r_j(t)$ where
\begin{equation}\label{1.2}
{v_i}\left( t \right) = \sum\nolimits_j {\int_{ - \infty }^t {{w_{ij}}{h_j}\left( {t - t'} \right){r_j}\left( t \right)} } dt'.
\end{equation}
To convert the equation system into and integral differential system we shall take derivative of~\ref{Eq1.1} yielding
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  \frac{{d{v_i}\left( t \right)}}{{dt}} &=& \sum\nolimits_{j,k} {\left( {{w_{ij}}{h_j}\left( t- {t_k} \right)\frac{d}{{dt}}H\left( {t - {t_k}} \right) + {w_{ij}}H\left( {t - {t_k}} \right)\frac{d}{{dt}}{h_j}\left( t- {t_k} \right)} \right)} \nonumber \\
   &=& \sum\nolimits_{j,k} {{w_{ij}}{h_j}\left( t- {t_k} \right)\delta \left( {t - {t_k}} \right)}  - \sum\nolimits_{j,k} {{w_{ij}}H\left( {t - {t_k}} \right)\frac{1}{{{\tau _j}}}{h_j}\left( t- {t_k} \right)}   \nonumber \\
   &=& \sum\nolimits_{j,k} {{w_{ij}}\delta \left( {t - {t_k}} \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\frac{1}{{{\tau _j}}}} \nonumber  \\
   &=& \sum\nolimits_j {{w_{ij}}{r_j}\left( t \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\frac{1}{{{\tau _j}}}} \nonumber \\
   &=& \sum\nolimits_j {{w_{ij}}P\left( {{v_j}\left( t \right)} \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\frac{1}{{{\tau _j}}}}.
\end{eqnarray}
This form can easily be extended to represent a continuous field.

\subsection{Assumptions and Simplifications}
We assume the velocity of the action potentials is infinite. This is also a common assumption that is made when modeling intracortical circuits. As stated in Wilson and Cohen's original works, the time constant $\tau$ is much greater than the axonal propagation delays, therefore it is reasonable to approximate the the delays as being infinite. Recent studies have demonstrated the importance of having a fast, but finite, propagation velocity for models to make realistic and complete predictions of cortical dynamics (Coombes paper, and Atay paper). The purpose of this paper, however, is not to precisely simulate complex cortical dynamics, but to create a useful platform where estimation of hidden states is possible for the purpose of detecting abnormal neural states with intentions of generating informed patient specific control sequences. Therefore, the model simply has to be 'good enough' for this purpose. Starting with this simplified version of the general model is a logical choice. The next assumption we will make is spatial homogeneity such that synaptic interaction between neural populations is only dependant on their spatial separation, and not on their spatial location. Again, this is a common assumption when dealing with intracortical connectivity. This assumption is not essential for our IDE estimation techniques, where a heterogeneous structure is possible. The IDE estimation is derived for a general heterogeneous structure, however for clarity of results we shall only present the heterogeneous case. We will assume the synaptic response kernel has the same shape (scaled by the connectivity kernel) irrespective of the respective presynaptic population. In this case inhibition can be modeled by having a negative response kernel and excitation with a positive kernel, where the kernel provides the (scaled) polarity. This reduced the model to a single field. I am not sure how this will limit the dynamics??
With these assumptions we have the model in the form
\begin{equation}\label{Eq10}
{v_{t + 1}}\left( s \right) = \int_\Omega  {W\left( {\left| {s - r} \right|} \right)P\left( {{v_t}\left( r \right)} \right)dy}  + {\varepsilon _t}\left( s \right) - \frac{1}{\tau }{v_t}\left( s \right).
\end{equation}

>>>>

\section{Neural Field Model}
In this section we describe a variant of the  Wilson and Cowan neural field model~\cite{Wilson1973} that will be used to create our connectivity estimator. This model has shown to be descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations and epileptic behavior. It is also capable of generating complex patterns of activity such as Turing patterns, spirals and traveling oscillations. The general form of the model is described here. The post-synaptic potentials generated at a neuronal population $r$ by action potentials arriving from all other connected populations at position $r'$ can be described by
\begin{equation}\label{SpikesToPotential}
v\left( {r,t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)f\left( {r,t'} \right)dt'} + e\left( {r,t} \right).
\end{equation}
Equation~\ref{SpikesToPotential} relates, $f$, or the average number of action potentials arriving at $r$ to the post synaptic potential at $r$, to the local membrane voltage. The term $e(r,t)$ can be considered the potential generated by unmodeled inputs and spontaneous background activity. The post-synaptic response kernel $h(t)$ is described by
\begin{equation}\label{SynapticRespKernel}
h(t) = \left\{ {\begin{array}{*{20}{c}}
   {\exp ( - \alpha t)} & {t \ge 0}  \\
   0 & {t < 0}  \\
\end{array}} \right.
\end{equation}
where $\alpha=\tau^{-1}$ and $\tau$ is the synaptic time constant. The synaptic response kernel can be more elaborate, using a rise and fall time constants, however this simple form serves our purposes. Nonlocal interactions between cortical populations are described by	
\begin{equation}\label{RateBasedInteractions}
f\left( {r,t} \right) = \int_\Omega  {w\left( {r,r'} \right)p\left( {v\left( {r',\bar t} \right)} \right)dr'}
\end{equation}
where $\bar t = t - d\left| {r - r'} \right|$, $p$ is spiking rate of populations $r'$, $w$ is the spatial connectivity kernel, $\Omega$ is the spatial domain, representing a cortical sheet or surface and $d$ is the propagation delay from $r'$ to $r$.

The average firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the Heaviside step function
\begin{equation}\label{RateOfPreSynPops}
p\left( {v\left( {r,\bar t} \right)} \right) = {\nu_0}u\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right)
\end{equation}
\begin{equation}\label{HeavisideDef}
u\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right) = \left\{ {\begin{array}{*{20}{c}}
   1 & {v\left( {r,\bar t} \right) \ge {v_0}\left( r \right)}  \\
   0 & {otherwise}  \\
\end{array}} \right.
\end{equation}
Even though we will model  $v_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality when constructing the state-space representation in the next section. The parameters in equation~\ref{RateOfPreSynPops} are... $v_0$ is typically 6~mV, $\nu_0$ is 5~$s^{-1}$ (from Wendling and David papers). By combining equations~\ref{SpikesToPotential} and \ref{RateBasedInteractions} we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
v\left(r,t\right) = \int\limits_{-\infty}^t\int\limits_\Omega  h\left(t - t'\right)w\left(r,r'\right)p\left( v\left( r',\bar t \right)\right)dr' dt'+e\left(r,t\right)
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{GenForm}
L\left( {\frac{\partial }{{dt}}} \right)v\left( {r,t} \right) = \int_\Omega  {w\left( {r,r'} \right)p\left( {v\left( {r',\bar t} \right)} \right)dr'}  + e\left( {r,t} \right),
\end{equation}
where $L$ is a polynomial of order $n$ with constant coefficients that provides an $n^{th}$ order model. Most studies considering this model use first or second order time derivatives. We shall use the first order model giving
\begin{equation}\label{FinalForm1}
\frac{{dv\left( {r,t} \right)}}{{dt}} + \alpha{v\left( {r,t} \right)} = \int_\Omega  {w\left( {r,r'} \right)p\left( {v\left( {r',\bar t} \right)} \right)dr'}  +e\left( {r,t} \right).
\end{equation}
\subsection{Alternate Model Description I}
An alternate and equivalent way to remove the temporal integration using the Laplace transform . We start by considering a discrete neural mass $i$ with inputs with zero delay from another discrete mass $j$ and
\begin{equation}\label{TwoDiscreteMasses}
{v_i}\left( t \right) = \int_{ - \infty }^t {{h_j}(t - t'){w_{ij}}p\left( {{v_j}\left( t \right)} \right)dt'},
\end{equation}
where $w_{ij}$ can be considered the synaptic efficiency from $j$ to $i$. Now we take the Laplace transform $\mathcal{L}(v_i(t))=V(s)$ yielding
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  {V_i}\left( s \right) &=& {H_j}\left( s \right){w_{ij}}P\left( {{V_j}\left( s \right)} \right) \nonumber \\
   &=& {\left( {s + \alpha} \right)^{ - 1}}{w_{ij}}P\left( {{V_j}\left( s \right)} \right)  \\
  \Rightarrow s{V_i}\left( s \right) + \alpha{V_i}\left( s \right) &=& {w_{ij}}P\left( {{V_j}\left( s \right)} \right).
\end{eqnarray}
Now taking the inverse Laplace transform we get
\begin{equation}\label{InverseLaplace}
\frac{{d{v_i}\left( t \right)}}{{dt}} + \alpha{v_i}\left( t \right) = {w_{ij}}p\left( {{v_j}\left( t \right)} \right).
\end{equation}
Next we convert the model from discrete neural masses to a continuum by
\begin{equation}\label{Discrete2Continuum}
\frac{{dv\left( {r,t} \right)}}{{dt}} + \alpha{v\left( {r,t} \right)} = \int_\Omega  {w\left( {r,r'} \right)p\left( {v\left( {r',\bar t} \right)} \right)dr'}  + e\left( {s,t} \right).
\end{equation}
The neural field model is popular due to being parsimonious yet having a strong link with the underlying physiology.

@@

An alternative description of the model is as follows. The model can be thought of as a continuum of neural masses. To illustrate this idea we shall first describe the interaction of discrete neural masses as seen in Figure~\ref{NeuralMassModel}. Assuming post-synaptic potentials sum linearly, the membrane potential at the soma of neuron $i$ is described as
\begin{equation}\label{VoltageFromPSPs}
{v_i}\left( t \right) = \sum\nolimits_{j,k} {{w_{ij}}{h_{j}}\left( t - {t_k}\right)} u\left( {t - {t_k}} \right),
\end{equation}
where $w_{ij}$ is the synaptic weight from neuron $j$ to $i$, $h(t)$ is the synaptic response kernel, $S$ is the Heaviside step function and action potential arrive at neuron $i$ at times $t_k$. Alternatively the membrane potential can described by the convolution of synaptic response kernel with the pre-synaptic firing rate $r_j(t)$ where
\begin{equation}\label{VoltageFromRate}
{v_i}\left( t \right) = \sum\nolimits_j {\int_{ - \infty }^t {{w_{ij}}{h_j}\left( {t - t'} \right){r_j}\left( t \right)} } dt'.
\end{equation}
To convert the equation system into and integral differential system we shall take derivative of~\ref{Eq1.1} yielding
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  \frac{{d{v_i}\left( t \right)}}{{dt}} &=& \sum\nolimits_{j,k} {\left( {{w_{ij}}{h_j}\left( t- {t_k} \right)\frac{d}{{dt}}u\left( {t - {t_k}} \right) + {w_{ij}}u\left( {t - {t_k}} \right)\frac{d}{{dt}}{h_j}\left( t- {t_k} \right)} \right)} \nonumber \\
   &=& \sum\nolimits_{j,k} {{w_{ij}}{h_j}\left( t- {t_k} \right)\delta \left( {t - {t_k}} \right)}  - \sum\nolimits_{j,k} {{w_{ij}}u\left( {t - {t_k}} \right)\alpha_j{h_j}\left( t- {t_k} \right)}   \nonumber \\
   &=& \sum\nolimits_{j,k} {{w_{ij}}\delta \left( {t - {t_k}} \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\alpha_j} \nonumber  \\
   &=& \sum\nolimits_j {{w_{ij}}{\nu_j}\left( t \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\alpha_j} \nonumber \\
   &=& \sum\nolimits_j {{w_{ij}}p\left( {{v_j}\left( t \right)} \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\alpha_j}.
\end{eqnarray}
This form can easily be extended to represent a continuous field.

\subsection{Assumptions and Simplifications}
We assume the velocity of the action potentials is infinite. This is also a common assumption that is made when modeling intracortical circuits. As stated in Wilson and Cohen's original works, the time constant $\tau$ is much greater than the axonal propagation delays, therefore it is reasonable to approximate the the delays as being infinite. Recent studies have demonstrated the importance of having a fast, but finite, propagation velocity for models to make realistic and complete predictions of cortical dynamics (Coombes paper, and Atay paper). The purpose of this paper, however, is not to precisely simulate complex cortical dynamics, but to create a useful platform where estimation of hidden states is possible for the purpose of detecting abnormal neural states with intentions of generating informed patient specific control sequences. Therefore, the model simply has to be 'good enough' for this purpose. Starting with this simplified version of the general model is a logical choice. The next assumption we will make is spatial homogeneity such that synaptic interaction between neural populations is only dependant on their spatial separation, and not on their spatial location. Again, this is a common assumption when dealing with intracortical connectivity. This assumption is not essential for our IDE estimation techniques, where a heterogeneous structure is possible. The IDE estimation is derived for a general heterogeneous structure, however for clarity of results we shall only present the heterogeneous case. We will assume the synaptic response kernel has the same shape (scaled by the connectivity kernel) irrespective of the respective presynaptic population. In this case inhibition can be modeled by having a negative response kernel and excitation with a positive kernel, where the kernel provides the (scaled) polarity. This reduced the model to a single field. I am not sure how this will limit the dynamics??
With these assumptions we have the model in the form
\begin{equation}\label{SimplifiedModel}
\frac{{dv\left( {s,t} \right)}}{{dt}} = \int_\Omega  {w\left( {s-r} \right)P\left( {v\left( {r,t} \right)} \right)dr}  - \alpha{v\left( {s,t} \right)} + e\left( {s,t} \right).
\end{equation}

\subsection{IDE Formulation}

\begin{equation}\label{new_spatiotemporal_kernels}
\frac{{dv\left( {s,t} \right)}}{{dt}} = \int_\Omega  {k\left( {s - r} \right)\tilde v\left( {r,t} \right) - \alpha \delta \left( {s - r} \right)\bar v\left( {s,t} \right)dr}  + e\left( {s,t} \right)
\end{equation}

\begin{equation}\label{define_v_tilde}
\tilde v\left( {r,t} \right) = \left\{ {\begin{array}{*{20}{c}}
   {v\left( {r,t} \right)} & {v\left( {r,t} \right) \ge {v_0}}  \\
   0 & {otherwise}  \\
\end{array}} \right.
\end{equation}

\begin{equation}\label{define_v_bar}
\bar v\left( {r,t} \right) = \left\{ {\begin{array}{*{20}{c}}
   {v\left( {r,t} \right)} & {v\left( {r,t} \right) < {v_0}}  \\
   0 & {otherwise}  \\
\end{array}} \right.
\end{equation}
